```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This workbook contains census data on different population demographics for Champaign, Illinois. I analyzed trends at the city, county, and regional levels to gain a better understanding of population dynamics and conditions in the area. This analysis was guided by instructions and examples from UP494: Neighborhood Analysis.


```{r}
library(blogdown)
library(tidycensus)
library(tidyverse)
```

First, I created an object for the population by race for all places in Illinois.

```{r error=FALSE, message=FALSE}
place_race_2017<-get_acs(geography = "place", state = "Illinois", table = "B02001", year=2017, survey="acs5", output="wide")

place_race_2017
```

Next, I repeated the process for counties in Illinois.

```{r error=FALSE, message=FALSE}
county_race_2017<-get_acs(geography = "county", state = "IL", table = "B02001", year=2017, survey="acs5", output="wide")

county_race_2017
```

Here, I altered the initial call for the data by including only the variables that were most relevant for creating a concise summary of racial demographics.

```{r error=FALSE, message=FALSE}
place_race_2017<-get_acs(geography = "place", state = "IL", variables = c("B02001_001", "B02001_002", "B02001_003"), year=2017, survey="acs5", output="wide")

place_race_2017
```

After filtering the data for Champaign, renaming variables and removing the rows I didn't need, I was left with a breakdown of Champaign's population by race.

```{r}
place_race_2017 %>% 
  filter(NAME == "Champaign city, Illinois") %>% 
  mutate(pop_other = B02001_001E-(B02001_002E + B02001_003E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_other)
```

This step rearranged the rows for the overall race breakdown by place into alphabetical order using the `arrange()` command.

```{r}
place_race_2017 %>% arrange(NAME)
```

Here, I used the `desc()` command to reverse the alphabetical order of row entries for the data.

```{r}
place_race_2017 %>% arrange(desc(NAME))
```

This step uses `str_detect()` as an alternative way to search for data on Champaign:
```{r}
place_race_2017 %>% filter(str_detect(NAME, "Champaign"))
```

This step presents an improved, organized version of the Champaign race data.
```{r}
place_race_2017 %>% 
  filter(NAME == "Champaign city, Illinois") %>% 
  mutate(pop_other = B02001_001E-(B02001_002E + B02001_003E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_other)
```

Because the population of the "pop_other" category was so high, I redownloaded the dataset to see if there were additional groups I could separate the data into.

```{r}
place_race_2017<-get_acs(geography = "place", state = "IL", table = "B02001", year=2017, survey="acs5", output="wide")

place_race_2017 %>% filter(NAME == "Champaign city, Illinois")
```

This lead to adding an additional category for Asian residents in Champaign based on their sizeable population.

```{r}
place_race_2017 %>% 
  filter(NAME == "Champaign city, Illinois") %>% 
  mutate(pop_other = B02001_001E-(B02001_002E + B02001_003E+ B02001_005E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E, pop_asian = B02001_005E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_asian, pop_other)

```

I then added this group into the Champaign race object.

```{r}
place_race_2017<-place_race_2017 %>% 
  filter(NAME == "Champaign city, Illinois") %>% 
  mutate(pop_other = B02001_001E-(B02001_002E + B02001_003E+ B02001_005E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E, pop_asian = B02001_005E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_asian, pop_other)
place_race_2017
```

After completing that process, I repeated it for Champaign County as well. 

```{r}
county_race_2017 <- county_race_2017 %>% 
  mutate(pop_other = B02001_001E-(B02001_002E+B02001_003E+B02001_005E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E, pop_asian = B02001_005E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_asian, pop_other)
county_race_2017 <- county_race_2017 %>% filter(str_detect(NAME, "Champaign"))
county_race_2017
```

I then worked on adding population percentages to both the place and county breakdowns.

```{r}
place_race_2017 <- place_race_2017 %>% 
  mutate(
    p_white = (pop_white/pop_tot)*100,
    p_black = (pop_black/pop_tot)*100,
    p_asian = (pop_asian/pop_tot)*100,
    p_other = (pop_other/pop_tot)*100
  )
```

```{r}
county_race_2017 <- county_race_2017 %>% 
  mutate(
    p_white = (pop_white/pop_tot)*100,
    p_black = (pop_black/pop_tot)*100,
    p_asian = (pop_asian/pop_tot)*100,
    p_other = (pop_other/pop_tot)*100
  )
```

After that, I bound both the place and county data into one object.

```{r}
race_2017<-bind_rows(place_race_2017, county_race_2017)
race_2017
```

The below chunk shows the population numbers for these geographic areas.

```{r}
race_2017 %>% select(NAME, pop_white, pop_black, pop_asian, pop_tot)
```

The below chunk shows the population percentages for these geographic areas.

```{r}
race_2017 %>% select(NAME, p_white, p_black, p_asian, p_other)
```

Next, I repeated the entire process again for the Champaign region, which includes the counties of Champaign, Ford, and Piatt.

```{r}
region_race_2017<- get_acs(geography = "county", state = "IL", table = "B02001", year=2017, survey="acs5", output="wide") %>% 
  filter(NAME %in% c("Champaign County, Illinois", "Ford County, Illinois", "Piatt County, Illinois")) %>% 
  summarise(B02001_001E = sum(B02001_001E),
            B02001_002E = sum(B02001_002E),
            B02001_003E = sum(B02001_003E),
            B02001_005E = sum(B02001_005E)) %>% 
  mutate(NAME = "Region")%>% 
  mutate(pop_other = B02001_001E-(B02001_002E + B02001_003E+ B02001_005E)) %>% 
  rename(pop_tot = B02001_001E, pop_white = B02001_002E, pop_black = B02001_003E, pop_asian = B02001_005E) %>% 
  select(NAME, pop_tot, pop_white, pop_black, pop_asian, pop_other) %>% 
  mutate(
  p_white = (pop_white / pop_tot)*100,
  p_black = (pop_black / pop_tot)*100,
  p_asian = (pop_asian / pop_tot)*100,
  p_other = (pop_other / pop_tot)*100)
region_race_2017
```

I also added the regional numbers into the combined city and county object.

```{r}
race_2017<-bind_rows(race_2017, region_race_2017)
race_2017
```

Next, I created a similar table that reflects Latino ethnicity for Champaign at the city level.

```{r}
place_latino_2017 <- get_acs(geography = "place", table = "B03002", year = 2017, state = "Illinois", survey = "acs5", output = "wide")

region_latino_2017 <- get_acs(geography = "county", table = "B03002", year = 2017, state = "IL", survey = "acs5", output = "wide")

place_latino_2017 <- place_latino_2017 %>% 
  filter(str_detect(NAME, "Champaign")) %>%
  mutate(latino_other = B03002_012E - (B03002_013E+B03002_014E)) %>% 
  rename(latino_tot = B03002_012E, latino_white = B03002_013E, latino_black = B03002_014E) %>% 
    select(NAME, latino_tot, latino_white, latino_black, latino_other)
place_latino_2017

region_latino_2017 <- region_latino_2017 %>% 
  filter(str_detect(NAME, "Champaign")) %>%
  mutate(latino_other = B03002_012E - (B03002_013E+B03002_014E)) %>% 
  rename(latino_tot = B03002_012E, latino_white = B03002_013E, latino_black = B03002_014E) %>% 
    select(NAME, latino_tot, latino_white, latino_black, latino_other)
place_latino_2017

latino_2017 <- bind_rows(place_latino_2017, region_latino_2017)
```

The next Census table I dealt with showed population by age.

```{r}
place_age_2017 <- get_acs(geography = "place", table = "B01001", year = 2017, state = "Illinois", survey = "acs5", output = "tidy")

place_age_2017
```

I first queried the data for Champaign.

```{r}
place_age_2017 <- place_age_2017 %>% 
  filter(str_detect(NAME, "Champaign"))
place_age_2017
```

Next, I used the `str_replace()` function to remove the summary rows from the data to make analysis easier. I also removed the table name from each entry, leaving only the specific indicator for each variable.

```{r}
place_age_2017<-place_age_2017 %>% 
  mutate(variable = str_replace(variable, "B01001_0", "")) %>% filter(!variable %in% c("01", "02", "26"))

place_age_2017
```

Next, I used the `case_when()` command to recode the data and make the variables more readable.

```{r}
?case_when
place_age_2017<-place_age_2017 %>% 
  mutate(variable_cat = case_when(
    variable == "03" ~ "19 and Below", 
    variable == "04" ~ "19 and Below", 
    variable == "05" ~ "19 and Below",
    variable == "06" ~ "19 and Below",
    variable == "07" ~ "19 and Below",
    variable == "08" ~ "20 to 44",
    variable == "09" ~ "20 to 44",
    variable == "10" ~ "20 to 44",
    variable == "11" ~ "20 to 44",
    variable == "12" ~ "20 to 44",
    variable == "13" ~ "20 to 44",
    variable == "14" ~ "20 to 44",
    variable == "15" ~ "45 to 64",
    variable == "16" ~ "45 to 64",
    variable == "17" ~ "45 to 64",
    variable == "18" ~ "45 to 64",
    variable == "19" ~ "65 and Above",
    variable == "20" ~ "65 and Above",
    variable == "21" ~ "65 and Above",
    variable == "22" ~ "65 and Above",
    variable == "23" ~ "65 and Above",
    variable == "24" ~ "65 and Above",
    variable == "25" ~ "65 and Above",
    variable == "27" ~ "19 and Below",
    variable == "28" ~ "19 and Below",
    variable == "29" ~ "19 and Below",
    variable == "30" ~ "19 and Below",
    variable == "31" ~ "19 and Below",
    variable == "32" ~ "20 to 44",
    variable == "33" ~ "20 to 44",
    variable == "34" ~ "20 to 44",
    variable == "35" ~ "20 to 44",
    variable == "36" ~ "20 to 44",
    variable == "37" ~ "20 to 44",
    variable == "38" ~ "20 to 44",
    variable == "39" ~ "45 to 64",
    variable == "40" ~ "45 to 64",
    variable == "41" ~ "45 to 64",
    variable == "42" ~ "45 to 64",
    variable == "43" ~ "45 to 64",
    variable == "44" ~ "65 and Above", 
    variable == "45" ~ "65 and Above",
    variable == "46" ~ "65 and Above",
    variable == "47" ~ "65 and Above",
    variable == "48" ~ "65 and Above",
    variable == "49" ~ "65 and Above"
    )
  )

place_age_2017
```

Next, I grouped the data by the recoded variables and summarized the values for each age category.

```{r}
place_age_2017<-place_age_2017 %>% 
  group_by(variable_cat) %>% 
  summarise(age=sum(estimate))

place_age_2017
```

I then repeated the process for the county and region as well.

```{r}
counties_age_2017 <- get_acs(geography = "county", table = "B01001", year = 2017, state = "Illinois", survey = "acs5", output = "tidy")

counties_age_2017 <- counties_age_2017 %>%
  filter(str_detect(NAME, "Champaign")) %>% 
  mutate(variable = str_replace(variable, "B01001_0", "")) %>% 
  filter(!variable %in% c("01", "02", "26")) %>%
  mutate(variable_cat = case_when(
    variable == "03" ~ "19 and Below", 
    variable == "04" ~ "19 and Below", 
    variable == "05" ~ "19 and Below",
    variable == "06" ~ "19 and Below",
    variable == "07" ~ "19 and Below",
    variable == "08" ~ "20 to 44",
    variable == "09" ~ "20 to 44",
    variable == "10" ~ "20 to 44",
    variable == "11" ~ "20 to 44",
    variable == "12" ~ "20 to 44",
    variable == "13" ~ "20 to 44",
    variable == "14" ~ "20 to 44",
    variable == "15" ~ "45 to 64",
    variable == "16" ~ "45 to 64",
    variable == "17" ~ "45 to 64",
    variable == "18" ~ "45 to 64",
    variable == "19" ~ "65 and Above",
    variable == "20" ~ "65 and Above",
    variable == "21" ~ "65 and Above",
    variable == "22" ~ "65 and Above",
    variable == "23" ~ "65 and Above",
    variable == "24" ~ "65 and Above",
    variable == "25" ~ "65 and Above",
    variable == "27" ~ "19 and Below",
    variable == "28" ~ "19 and Below",
    variable == "29" ~ "19 and Below",
    variable == "30" ~ "19 and Below",
    variable == "31" ~ "19 and Below",
    variable == "32" ~ "20 to 44",
    variable == "33" ~ "20 to 44",
    variable == "34" ~ "20 to 44",
    variable == "35" ~ "20 to 44",
    variable == "36" ~ "20 to 44",
    variable == "37" ~ "20 to 44",
    variable == "38" ~ "20 to 44",
    variable == "39" ~ "45 to 64",
    variable == "40" ~ "45 to 64",
    variable == "41" ~ "45 to 64",
    variable == "42" ~ "45 to 64",
    variable == "43" ~ "45 to 64",
    variable == "44" ~ "65 and Above", 
    variable == "45" ~ "65 and Above",
    variable == "46" ~ "65 and Above",
    variable == "47" ~ "65 and Above",
    variable == "48" ~ "65 and Above",
    variable == "49" ~ "65 and Above")
    ) %>% 
  group_by(variable_cat) %>% 
  summarise(age=sum(estimate))
  
counties_age_2017

```

```{r}
region_age_2017 <- get_acs(geography = "county", table = "B01001", year = 2017, state = "Illinois", survey = "acs5", output = "tidy" ) %>% 
  filter(NAME %in% c("Champaign County, Illinois", "Ford County, Illinois", "Piatt County, Illinois")) %>% 
  mutate(variable = str_replace(variable, "B01001_0", "")) %>% 
  filter(!variable %in% c("01", "02", "26")) %>%
  mutate(variable_cat = case_when(
    variable == "03" ~ "19 and Below", 
    variable == "04" ~ "19 and Below", 
    variable == "05" ~ "19 and Below",
    variable == "06" ~ "19 and Below",
    variable == "07" ~ "19 and Below",
    variable == "08" ~ "20 to 44",
    variable == "09" ~ "20 to 44",
    variable == "10" ~ "20 to 44",
    variable == "11" ~ "20 to 44",
    variable == "12" ~ "20 to 44",
    variable == "13" ~ "20 to 44",
    variable == "14" ~ "20 to 44",
    variable == "15" ~ "45 to 64",
    variable == "16" ~ "45 to 64",
    variable == "17" ~ "45 to 64",
    variable == "18" ~ "45 to 64",
    variable == "19" ~ "65 and Above",
    variable == "20" ~ "65 and Above",
    variable == "21" ~ "65 and Above",
    variable == "22" ~ "65 and Above",
    variable == "23" ~ "65 and Above",
    variable == "24" ~ "65 and Above",
    variable == "25" ~ "65 and Above",
    variable == "27" ~ "19 and Below",
    variable == "28" ~ "19 and Below",
    variable == "29" ~ "19 and Below",
    variable == "30" ~ "19 and Below",
    variable == "31" ~ "19 and Below",
    variable == "32" ~ "20 to 44",
    variable == "33" ~ "20 to 44",
    variable == "34" ~ "20 to 44",
    variable == "35" ~ "20 to 44",
    variable == "36" ~ "20 to 44",
    variable == "37" ~ "20 to 44",
    variable == "38" ~ "20 to 44",
    variable == "39" ~ "45 to 64",
    variable == "40" ~ "45 to 64",
    variable == "41" ~ "45 to 64",
    variable == "42" ~ "45 to 64",
    variable == "43" ~ "45 to 64",
    variable == "44" ~ "65 and Above", 
    variable == "45" ~ "65 and Above",
    variable == "46" ~ "65 and Above",
    variable == "47" ~ "65 and Above",
    variable == "48" ~ "65 and Above",
    variable == "49" ~ "65 and Above")
    ) %>% 
  group_by(variable_cat) %>% 
  summarise(age=sum(estimate))

region_age_2017

```

Next, I used `bind_cols()`to combine the tables.

```{r}
bind_cols(place_age_2017, counties_age_2017, region_age_2017)
```

The `select()` command allowed me to choose only the variables that I wanted to include in the table.

```{r}
age_2017<-bind_cols(place_age_2017, counties_age_2017, region_age_2017) %>% 
  select(Age = variable_cat...1, Place = age...2, County = age...4, Region = age...6) 

age_2017
```

Next, I calculated percentages for each geographical area.

```{r}
age_2017<-age_2017 %>% 
  mutate(p_Place = (Place / sum(Place))*100,
         p_County = (County / sum(County))*100,
         p_Region = (Region / sum(Region))*100,
         )

age_2017
```

After completing the race and age tables, I decided to take a look at poverty in Champaign and repeat the analysis for that data.

The below table shows poverty for the City of Champaign.

```{r}

place_poverty_2017 <- get_acs(geography = "place", state = "Illinois", table = "B17001", year = 2017, survey = "acs5", output = "wide")

place_poverty_2017 <- place_poverty_2017 %>% 
  filter(str_detect(NAME, "Champaign")) %>% 
  mutate(
    tot_pop = B17001_001E,
    tot_below_poverty = B17001_002E,
    male_below_poverty = B17001_003E,
    female_below_poverty = B17001_017E,
    tot_above_poverty = B17001_031E,
    male_above_poverty = B17001_032E,
    female_above_poverty = B17001_046E
  ) %>% 
  select(NAME, tot_pop, tot_below_poverty, male_below_poverty, female_below_poverty, tot_above_poverty, male_above_poverty, female_above_poverty)
  
place_poverty_2017 <- place_poverty_2017 %>% 
  mutate(
    p_below_poverty = (tot_below_poverty/tot_pop)*100,
    p_above_poverty = (tot_above_poverty/tot_pop)*100,
    p_male_below_poverty = (male_below_poverty/tot_below_poverty),
    p_female_below_poverty = (female_below_poverty/tot_below_poverty)*100,
    p_male_above_poverty = (male_above_poverty/tot_above_poverty)*100,
    p_female_above_poverty = (female_above_poverty/tot_above_poverty)*100
  )

place_poverty_2017
  
```

The below table shows poverty for Champaign County.

```{r}
county_poverty_2017 <- get_acs(geography = "county", state = "Illinois", table = "B17001", year = 2017, survey = "acs5", output = "wide")

county_poverty_2017 <- county_poverty_2017 %>% 
  filter(str_detect(NAME, "Champaign")) %>% 
  mutate(
    tot_pop = B17001_001E,
    tot_below_poverty = B17001_002E,
    male_below_poverty = B17001_003E,
    female_below_poverty = B17001_017E,
    tot_above_poverty = B17001_031E,
    male_above_poverty = B17001_032E,
    female_above_poverty = B17001_046E
  ) %>% 
  select(NAME, tot_pop, tot_below_poverty, male_below_poverty, female_below_poverty, tot_above_poverty, male_above_poverty, female_above_poverty)
  
county_poverty_2017 <- county_poverty_2017 %>% 
  mutate(
    p_below_poverty = (tot_below_poverty/tot_pop)*100,
    p_above_poverty = (tot_above_poverty/tot_pop)*100,
    p_male_below_poverty = (male_below_poverty/tot_below_poverty),
    p_female_below_poverty = (female_below_poverty/tot_below_poverty)*100,
    p_male_above_poverty = (male_above_poverty/tot_above_poverty)*100,
    p_female_above_poverty = (female_above_poverty/tot_above_poverty)*100
  )

county_poverty_2017
```

The below table shows poverty for the Champaign region.

```{r}
region_poverty_2017 <- get_acs(geography = "county", state = "Illinois", table = "B17001", year = 2017, survey = "acs5", output = "wide")

region_poverty_2017 <- region_poverty_2017 %>% 
  filter(NAME %in% c("Champaign County, Illinois", "Ford County, Illinois", "Piatt County, Illinois")) %>% 
  summarise(
    B17001_001E = sum(B17001_001E),
    B17001_002E = sum(B17001_002E),
    B17001_003E = sum(B17001_003E),
    B17001_017E = sum(B17001_017E),
    B17001_031E = sum(B17001_031E),
    B17001_032E = sum(B17001_032E),
    B17001_046E = sum(B17001_046E)
  ) %>% 
  mutate(
    NAME = "Champaign Region",
    tot_pop = B17001_001E,
    tot_below_poverty = B17001_002E,
    male_below_poverty = B17001_003E,
    female_below_poverty = B17001_017E,
    tot_above_poverty = B17001_031E,
    male_above_poverty = B17001_032E,
    female_above_poverty = B17001_046E
  ) %>% 
  select(NAME, tot_pop, tot_below_poverty, male_below_poverty, female_below_poverty, tot_above_poverty, male_above_poverty, female_above_poverty)
  
region_poverty_2017 <- region_poverty_2017 %>% 
  mutate(
    p_below_poverty = (tot_below_poverty/tot_pop)*100,
    p_above_poverty = (tot_above_poverty/tot_pop)*100,
    p_male_below_poverty = (male_below_poverty/tot_below_poverty),
    p_female_below_poverty = (female_below_poverty/tot_below_poverty)*100,
    p_male_above_poverty = (male_above_poverty/tot_above_poverty)*100,
    p_female_above_poverty = (female_above_poverty/tot_above_poverty)*100
  )

region_poverty_2017
```

Here, I bound the individual tables into one summary table. Poverty numbers are shown first, then percentages.

```{r}
champaign_poverty_2017 <- bind_rows(place_poverty_2017, county_poverty_2017, region_poverty_2017)

champaign_poverty_2017
```

### What do these indicators help you to say about how the qualities of Champaign, Champaign County, and the Region?

Combining data on the demographic characteristics of three different geographic extents allows us to understand how Champaign demographics compare with the surrounding areas. The race tables show that Champaign has a slightly higher number of minority residents than the surrounding region, with 34% non-white residents versus 27% non-white and 24% non-white at the county and regional levels respectively. From the age tables, the city of Champaign also seems to have a higher population of residents between the ages of 20 to 44 than the county and region, which makes sense with the University of Illinois's location there. The city has a slightly lower population of residents aged 45 to 64 than the county and region as well. The city, county, and region all have between 60% to 65% of their population between 20 to 64 years old. 
